---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
Load the knitr package and set the global options for the code chunks
```{r echo=TRUE}
library(knitr)
opts_chunk$set(fig.width=12, fig.height=8, fig.path='figure/',echo=TRUE)
```


Read the activity data

```{r}
data <- read.csv("activity.csv",stringsAsFactors = FALSE)
str(data)
```

There are 1440 (24*60) minutes in a day. The data has been taken at every 5 minutes interval for the month of October and November in 2012. Check that there are 288 entries (1440/5) for each date of the two months.

```{r}
table(data$date)
```

Conver the character date format into correct date format in the data frame
```{r}
data$date = as.Date(data$date,format("%Y-%m-%d"))
```
Check data has been properly transformed to reflect dates correctly
```{r}
str(data)
```
## What is mean total number of steps taken per day?



## What is the average daily activity pattern?



## Imputing missing values



## Are there differences in activity patterns between weekdays and weekends?
