---
title: "Reproducible Research: Peer Assessment 1"
author: Rajeev Trikha
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
Load the knitr package and set the global options for the code chunks
```{r echo=TRUE}
library(knitr)
opts_chunk$set(fig.width=12, fig.height=8, fig.path='figure/',echo=TRUE)
```


Read the activity data from the working directory

```{r}
data <- read.csv("activity.csv",stringsAsFactors = FALSE)
str(data)
```

There are 1440 (24*60) minutes in a day. The data has been taken at every 5 minutes interval for the month of October and November in 2012. Check that there are 288 entries (1440/5) for each date of the two months.

```{r}
table(data$date)
```

Conver the character date format into correct date format in the data frame
```{r}
data$date = as.Date(data$date,format("%Y-%m-%d"))
```
Check data has been properly transformed to reflect dates correctly
```{r}
str(data)
```
## What is mean total number of steps taken per day?

Establish total number of steps taken each day, whilst ignoring NAs

```{r}
library(dplyr)
dailyTotalSteps <- group_by(data,date) %>%
                            summarise(sum(steps, na.rm=TRUE))
```

Plot the histogram of total steps taken each day using 15 breaks and show mean and median steps.
```{r totalStepsPlot}
names(dailyTotalSteps)[2] <- "steps"
daily_mean <- round(mean(dailyTotalSteps$steps),0)
daily_median <- round(median(dailyTotalSteps$steps),0)
hist(dailyTotalSteps$steps,breaks=15,main="Number of steps taken per day",xlab="Total steps",col="grey")
abline(v=daily_mean, lwd = 2, col = "blue")
abline(v=daily_median, lwd = 2, col = "red")
legend("topright",fill=c("blue","red"),legend=c("mean","median"))
```

The mean value is `r daily_mean` and the median value is `r daily_median`. Check using 5 point summary
```{r}
summary(dailyTotalSteps$steps)
```

## What is the average daily activity pattern?

Establish average number of steps taken during each interval, whilst ignoring NAs

```{r}
stepsByInterval <- group_by(data,interval) %>%
                            summarise(mean(steps, na.rm=TRUE))
str(stepsByInterval)
```

Plot the average daily activity pattern

```{r averageSteps}
names(stepsByInterval)[2] <- "steps"
# Find maximum interval row and the values
maxrow <- which.max(stepsByInterval$steps)
maxInterval <- stepsByInterval[maxrow,1]
maxSteps <- stepsByInterval[maxrow,2]
#plot lines graph
plot(stepsByInterval$interval,stepsByInterval$steps,main="Average steps per interval",xlab="5 minute Intervals",ylab="Average Steps",pch="")
lines(stepsByInterval$interval,stepsByInterval$steps)
```

The maximum average steps of `r maxSteps` occurs in the interval `r maxInterval`.

## Imputing missing values

```{r}
missing_steps <- sum(is.na(data$steps))
missing_anywhere <- nrow(data) - sum(complete.cases(data))
```

Rows with NA in steps are `r missing_steps`. Number of rows with NA in at least one column are `r missing_anywhere`. As both values are equal, the NAs only exist in column steps.

We are going to assign the average value for the interval to the missing values.

```{r totsNAreplaced}
# use the stepsByInterval dataset created earlier for average steps and rename the seond column
names(stepsByInterval)[2] <- "avg_steps"
# Create a combined dataset which attaces the avg_steps column to the original data
new_data <- left_join(data,stepsByInterval) 
# replace NAs with average value
new_data$steps[which(is.na(new_data$steps))] <- new_data$avg_steps[which(is.na(new_data$steps))]

# calculate mean,median and plot
dailyTotalSteps <- group_by(new_data,date) %>%
                            summarise(sum(steps))
# Below is copied from the earlier plot but would normally be better off in a function
names(dailyTotalSteps)[2] <- "steps"
new_mean <- round(mean(dailyTotalSteps$steps),0)
new_median <- round(median(dailyTotalSteps$steps),0)
hist(dailyTotalSteps$steps,breaks=15,main="Number of steps taken per day with NAs replaced",xlab="Total steps",col="grey")
abline(v=new_mean, lwd = 2, col = "blue")
abline(v=new_median, lwd = 2, col = "red")
legend("topright",fill=c("blue","red"),legend=c("mean","median"))

```
The new mean value with NAs replaced is `r new_mean` and the median value is `r new_median`. Check using 5 point summary
```{r}
summary(dailyTotalSteps$steps)
```

The original values for the meaan and median were `r daily_mean` and `r daily_median`. It is apparent that the new mean and median are nearly the same. The mean and median lines in the plot are overlapping. The values for both the mean and median have increased too. It was a left skewed plot so replacing NAs has had a slight rightward shift.


## Are there differences in activity patterns between weekdays and weekends?
